Jane, John … Leslie? A Historical Method for Algorithmic Gender Prediction

### 2015-07-30 article.Rmd
Mostly AHR section

The "Leslie problem" is not just for researchers who think of themselves as historians. Anyone studying a period longer than a few years, or anyone studying a group whose demographics do not match the groups used by a contemporary tool such as Genderize.io will also encounter this problem. For example, a literary scholar studying a corpus of poems from the mid-twentieth century might wish to compare the stanza structures of male and female poets. Using a tool like Genderize.io based on contemporary data risks misidentifying these writers, many of who were born (and named) more than a century prior to the creation of these modern name datasets. This problem will only increase with growing life expectancies: as of 2012, the average American lifespan was nearly seventy-nine years---more than enough time for naming practices to change quite dramatically between a person’s birth and death [World Bank 2014]. Predicting gender from first names therefore requires a dynamic method that takes into change over time.

@@

Examples like the Pulitzer and Bancroft Prizes are only the most glaring cases of gender inequity in the historical profession. Overt discrimination still exists, but inequity operates in ways that are far more cloaked and far more complex, which frequently makes them difficult to recognize or remedy. They take the form of subconscious bias that subtly weakens the assessment of work done by female historians. Gender discrimination frequently intersects with other kinds of racial and socioeconomic inequity [Townsend 2008]. And larger societal structures impede the professional lives of women: a 2011 survey by the American Historical Association revealed that female historians dedicated substantially more time to child and elder care than their male colleagues, leaving them with less time for research [Townsend 2012]. All of these factors combine and amplify each other to contribute to the gender gap in the historical profession on display in the pages of the *American Historical Review*.

The hidden, murky, and systemic dimensions of gender inequity makes methods like the Gender package all the more imperative. Our program was able to infer the genders of tens of thousands of authors, aggregate and chart their trajectory over time, and compare the pattern to another dataset of dissertation authors. Doing so revealed an unmistakable gap in the number of women who were getting reviewed by teh *American Historical Review*. But our program was able to look at subtler patterns - the kind that make gender inequity so insidious. After inferring the gender of book authors reviewed by the *AHR*, we were also able to infer the gender of the reviewers themselves. The story is much the same for reviewers as it is for authors: more than twice as many men as women appear as reviewers in the journal. But gender also inflects book reviews by shaping who writes reviews of which authors. About three times as many men write reviews of male-authored books as do women. In the case of female-authored books the ratio of female-to-male reviewers is closer to 50/50. In short, women are far more likely to write reviews of other women, while men tend to gravitate more towards male authors – who are, of course, already over-represented in the *AHR*.

@@

Gender inequity is not always an overt case of deciding to award a prestigious prize to a male historian; it can be as subtle as deciding to write a book review of a male author. Countless individual actions like these constitute the relationships that knit together the field of academic history. And too often these tiny actions benefit men more than women: chairing a panel of mostly male historians, taking on a new male advisee, or assigning mostly male-authored books on a syllabus. On their own they might appear benign - one might argue that male historians simply happen to be interested in similar research topics - but in aggregate they contribute to a broader professional landscape that puts female historians at a disadvantage.

One 2013 analysis of 2,500 recent history Ph.D.'s found that "gender played little role in employment patterns across particular professions and industries" [Wood and Townsend 2013]. Our own analysis reveals a different story in terms of getting a book reviewed in the field's most prestigious journal. The *American Historical Review* continues to publish twice as many reviews of male-authored books as female-authored books. This disparity is put into even sharper relief when set against the much smaller gender gap in history dissertations. Women might be getting hired at comparable rates to men, but discriminatory gatekeeping remains alive and well in the profession.

@@

Include in conclusion
- gender is a tool that humanities scholars can put all of their tremendous skills in contextualizing, reocgnizing subtlety, and putting it in conversation with other sturctural power of race class and sexuality.
- Point to examples of how much it's downloaded, Ted Underwood, and especially Ben Schmidt's student evaluations, which is exactly the kind of ways it can be used to shed light on hidden patterns of systemic bias and discrimination.

>>>>

The "Leslie problem" is not just for researchers who think of themselves as historians. Anyone studying a period longer than a few years, or anyone studying a group whose demographics do not match the groups used by a contemporary tool such as Genderize.io will also encounter this problem. For example, a literary scholar studying a corpus of poems from the mid-twentieth century might wish to compare the stanza structures of male and female poets. Using a tool like Genderize.io based on contemporary data risks misidentifying these writers, many of whom were born (and named) more than a century prior to the creation of these modern name datasets. This problem will only increase with growing life expectancies: as of 2012, the average American lifespan was nearly seventy-nine years---more than enough time for naming practices to change quite dramatically between a person’s birth and death [World Bank 2014]. Predicting gender from first names therefore requires a dynamic method that takes into change over time.

@@

Examples like the Pulitzer and Bancroft Prizes are only the most glaring cases of gender inequity in the historical profession. Overt discrimination still exists, but inequity operates in ways that are far more cloaked and far more complex, making them difficult to recognize and remedy. They take the form of subconscious bias that subtly weakens the assessment of work done by female historians, whether in a seminar paper or a job application. Moreover, gender discrimination frequently intersects with other kinds of racial and socioeconomic inequity that erects even higher barriers in the women's careers [Townsend 2008]. Gendered family dynamics add another layer of challenges for women: a 2011 survey by the American Historical Association revealed that female historians dedicated substantially more time to child and elder care than their male colleagues, leaving them with less time for the kind of academic research that the profession rewards most highly [Townsend 2012]. All of these factors combine and amplify each other to contribute to the gender gap in the historical profession on display in the pages of the *American Historical Review*. The hidden, murky, and systemic dimensions of gender inequity makes methods like the Gender package all the more imperative for reasearchers.

Finally, we used the Gender package to reveal some of the subtler ways in which gender iniquity operates in the historical profession. After inferring the gender of book authors reviewed by the *AHR*, we were also able to infer the gender of the reviewers themselves. This led us to two findings. First, the story is much the same for reviewers as it is for authors: more than twice as many men as women appear as reviewers in the journal. This is important because reviewers for the *AHR* garner a certain degree of prestige from their work - perhaps not to the same prestige as winning a Pulitzer Prize, but prestigious nonetheless. To become a reviewer, one must possess a Ph.D. or equivalent degree and have published at least one book-length manuscript [American Historical Review n.d.]. *AHR* reviewers are understood to have sufficient experience and command over a given sub-field to evaluate the merits and weaknesses of a new contribution to that literature. Being a reviewer for the journal is, in many ways, a proxy for expertise. Which is a problem. When there are twice as many male "experts" as female "experts" appearing in the *AHR* it imbues our understanding of what constitutes historical expertise.  Or perhaps more accurately, who is and who isn't an "expert" in the field.

The second finding we arrived at when examining the gender of *AHR* reviewers was a better understanding of who reviews which book authors. Once again, the methodology reveals patterns that we might suspect but nonetheless tends to stay hidden from direct sight. About three times as many men as women write reviews of male-authored books in the *AHR*. In the case of female-authored books, the ratio of female-to-male reviewers is closer to 50/50. In short, women are far more likely to write reviews of other women, while men tend to gravitate more towards male authors – who are, of course, already over-represented in the *AHR*.

@@

The findings related to *AHR* reviewers help illustrate the ways in which gender inequity is so insidious. It is not always an overt case of deciding to award a prestigious prize to a male historian; it can be as seemingly innocuous as deciding to write a book review of a male author's book. Countless individual actions like these constitute the core relationships that knit together the profession, and too often these tiny actions benefit men more than women: chairing a panel of mostly male historians, taking on a new male advisee, or assigning mostly male-authored books on a syllabus. On their own they might appear benign - one might argue that male historians simply happen to be interested in similar research topics - but in aggregate they contribute to a broader professional landscape that puts female historians at a disadvantage.

One 2013 analysis of 2,500 recent history Ph.D.'s found that "gender played little role in employment patterns across particular professions and industries" [Wood and Townsend 2013]. Our own analysis reveals a different story, at least in terms of getting a book reviewed in the field's most prestigious journal. The *American Historical Review* continues to publish twice as many reviews of male-authored books as female-authored books, and those same reviews are more than twice as likely to be written by a man than a woman. This disparity is put into even sharper relief when set against the much smaller gender gap between female and male historians who complete dissertations. Women might be getting hired at comparable rates to men, but discriminatory gatekeeping remains alive and well in the profession. The Gender package is one more tool to uncover those practices.

@@

Include in conclusion
- gender is a tool that humanities scholars can put all of their tremendous skills in contextualizing, reocgnizing subtlety, and putting it in conversation with other sturctural power of race class and sexuality.
- Point to examples of how much it's downloaded, Ted Underwood, and especially Ben Schmidt's student evaluations, which is exactly the kind of ways it can be used to shed light on hidden patterns of systemic bias and discrimination.
- Posner's latest blog post on impreative of creating humanistic tools
